# Exercise 5. Analysis of longitudinal data

Load some libraries 
```{r}
library(dplyr) 
library(ggplot2) 
library(GGally)
library(tidyr)
```


First load the wrangled (longitudinal = L) datasets and original datasets into R 
```{r}
BPRSL <- read.table("C:/Users/JT/Documents/IODS/IODS-project/data/BPRSL.txt", 
                 sep = ",",
                 header = TRUE
                 )

RATSL <- read.table("C:/Users/JT/Documents/IODS/IODS-project/data/RATSL.txt", 
                 sep = ",",
                 header = TRUE
                 )

BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", 
                    sep = "",
                    header = TRUE
)

RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", 
                   sep = "\t",
                   header = TRUE
)

```


The RATSL dataset is from a rat nutrition study by Crowder and Hand (1990), in which three groups of rats were put on diﬀerent diets, and each animal’s body weight (grams) was recorded repeatedly (approximately
weekly, except in week seven when two recordings were taken) over a 9-week period.



The BPRSL dataset is from Davis (2002). In the experiment 40 male subjects were
randomly assigned to one of two treatment groups and each subject was rated
on the brief psychiatric rating scale (BPRS) measured before treatment be-
gan (week 0) and then at weekly intervals for eight weeks. The BPRS assesses
the level of 18 symptom constructs such as hostility, suspiciousness, halluci-
nations and grandiosity; each of these is rated from one (not present) to seven
(extremely severe). The scale is used to evaluate patients suspected of having
schizophrenia.


## Prepare the datasets
Now, let's check the data
```{r}

#The BPRSL dataset
names(BPRSL)
str(BPRSL)
summary(BPRSL)
glimpse(BPRSL)


#The RATSL dataset 
names(RATSL)
str(RATSL)
summary(RATSL)
glimpse(RATSL)
```

As it was observed that R did not load the factorized (categorial) variables correctly, it is necessary to re-factorize them 

```{r}
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
```

Ensure that the datasets looks as it should:
```{r}
glimpse(BPRSL)
glimpse(RATSL)
```

## Task 1: Implement the analyses of Chapter 8 of MABS using the RATS data

To start let's plot the raw data 
```{r}
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group))+scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))+scale_y_continuous(name = "Weight (grams)")+theme(legend.position = "top")
```


There are some clearly noticeable and systematic differences between the weights the rats in the different groups that are fed differently.

Now we plot mean response profiles for weight data the three groups of rats to inspect the behaviour of the mean weights
```{r}




# Number of days (Time), baseline (Time = 0) included
n <- RATSL$Time %>% unique() %>% length() 

# Summary data with mean and standard error of RATSL by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
 All mean weights of the groups (1-3) increased with time. Group 2 seemingly grew with the fastest rate. There was clearly more deviation within groups 2 and group 3 than within group 1. 
The groups differ from each other with at least one standard error at all time. I.e. there is clear difference between the groups

Now we plot a box plot of the data. Here we also filter out the baseline observations/measurements. Baseline measurements of the outcome variable in a longitudinal study are often correlated with the chosen summary measure and using such measures in the analysis can often lead to substantial gains in precision when used appropriately as a covariate in an analysis of covariance. 
```{r}
# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline i.e. Time 0). 
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)


# Draw a boxplot
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-9")

```
As interpreted earlier group 1 has smaller mean weight than group 2 and group 3. The difference between group 2 and group 3 is smaller, but also clear. 

The plot revealed some outliers (Weight < 250 g and weight > 550g). Let's filter theses out and re-draw the boxplot.
```{r}

# Filter the data
RATSL8S1 <- RATSL8S %>% filter(mean > 250 & mean < 550) 


glimpse(RATSL8S1)

# Draw the boxplot
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-9")
```
Removing the outlier makes the differences between the groups even more evident. 


Now a formal testing of difference will be done to examine the difference between the groups. The data with the excluded outliers and baseline is used. 
```{r}
# Between groups 1 and 2
t.test(mean ~ Group, data = filter(RATSL8S1, Group %in% c(1, 2)), var.equal=TRUE) 


# Between groups 2 and 3
t.test(mean ~ Group, data = filter(RATSL8S1, Group %in% c(2, 3)), var.equal=TRUE)

# Between groups 1 and 3
t.test(mean ~ Group, data = filter(RATSL8S1, Group %in% c(1, 3)), var.equal=FALSE)
```

The t-test result tables reveal that the mean weights of the groups differ from each other statistically significantly (p<0.05).


Finally the analysis of variance is. For purpose the baseline is added 
```{r}
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>% mutate(baseline =  RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data=RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

The anova summary reveals that the baseline weight relates stronlgy to the weights recorded after following the group specific diet. THere are no cler statistical significance with other variables. 


## Task 2: Implement the analyses of Chapter 9 of MABS using the BPRS data
First the data is plotted
```{r}
# Plot the RATSL data
# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```
  The above figure displays the behaviour of bprs of two treatments over time. On average bprs seemingly decreases over time. for both treatments. 



Next a linear regression model is created
```{r}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs ~ treatment + week, data = BPRSL)

# And print the summary of the model
summary(BPRS_reg)
```
The t-test part in the summary reveals that there is statistically significant relationship between the variable week (time) and the response variable bprs. Significant relationship between the treatment variable and bprs does not exist. 

Now, we will create a random intercept model to the data
```{r}
# access library lme4 necessary here
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
# Print the summary of the model
summary(BPRS_ref)

```

From the summary one can interpret that the t-value for treatment2 is very small, which indicates there is no statistically significant difference in the intercepts for observations with different treatments. The week variable has large negative t-value which indicates that the response variable decreases as week increases.

Next, create a random intercept and random slope model
```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref1)
```


Again the t-value for treatment2 is small. There is no statistically significant difference beteen the ficed effects of the two treatments



Now we create a random intercept and random slope model with the interaction,
```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = TRUE)

# print a summary of the model
summary(BPRS_ref2)
```

```{r}
# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

Finally draw the plots 
```{r}
# draw the plot of RATSL with the observed bprs values
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))




# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to RATSL
BPRSL <- BPRSL %>% mutate(BPRSL, Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```
